---
title: "Agentic workflow: closing the digest loop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Agentic workflow: closing the digest loop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

capR's core is a deterministic pipeline: `cap_digest()` compiles an object
into bounded evidence, `cap_validate_response()` checks a model's citations,
`cap_gate()` authorizes follow-up, and `cap_patch()`/`cap_apply_patch()`
extend the digest. The quickstart documents that round trip as a ten-step
recipe; the *agent session* layer ships it as code, so a model can actually
drive the loop while every disclosure still passes through the gate.

What this layer never does:

- It never calls a model provider. The model call is injected by the host
  (`ask` below) or lives in an external framework such as aisdk.
- It never bypasses the gate. Every tool call and every turn is synthesized
  into a `cap.contract_response.v1` and pushed through the full
  validate -> gate -> patch -> apply path.
- It never invents nondeterminism. Transcripts carry no timestamps and no
  random ids; identical runs serialize to identical canonical JSON.

## A closed loop with a scripted model

Any `function(text) -> response` can stand in for the model. Here a scripted
stand-in first requests the sample field, then answers with a claim that
cites it.

```{r}
library(capR)

orders <- data.frame(
  order_id = c("A001", "A002"),
  amount = c(12.5, 19)
)

session <- cap_agent_session(
  orders,
  question = "How many orders are there?",
  budget = 500,
  policy = cap_policy(max_followup_budget = 300)
)
session
```

The session eagerly builds the turn-0 digest. `cap_agent_prompt()` composes
the model-facing text: deterministic instructions plus the digest.

```{r}
cat(substr(cap_agent_prompt(session), 1L, 400L))
```

```{r}
scripted_model <- local({
  turn <- 0L
  function(prompt) {
    turn <<- turn + 1L
    if (turn == 1L) {
      list(
        claims = list(),
        evidence = list(),
        warnings = list(),
        requests = list(list(
          fieldId = "f1:table@sample#k10",
          level = 1L,
          budget = 300L,
          reason = "Need bounded sample rows."
        ))
      )
    } else {
      list(
        claims = list(list(
          id = "claim-1",
          text = "The table holds two orders.",
          evidence = list("f1:table@shape#base", "f1:table@sample#k10")
        )),
        evidence = list(),
        warnings = list(),
        requests = list()
      )
    }
  }
})

cap_agent_run(session, scripted_model)
session
```

Turn one was gated, patched, and applied; turn two answered from disclosed
evidence, closing the loop.

```{r}
vapply(session$turns, `[[`, character(1), "outcome")
session$followup_remaining
stopifnot(
  session$stop_reason == "completed",
  grepl('<field id="f1:table@sample#k10"', session$digest$text, fixed = TRUE)
)
```

The transcript is the audit trail: every turn embeds the canonical
validation, gate, and patch artifacts plus content hashes of the exact
prompt and response.

```{r}
transcript <- cap_agent_transcript(session)
transcript$finalBudget
transcript$turns[[1L]]$gate$overallDecision
```

Set `artifact_dir` to publish `turn-NNN/` artifact directories and an
atomically rewritten `transcript.capr.json` after every turn -- the same
canonical files `cap_write_artifacts()` produces elsewhere.

## Fail-closed outcomes

Denials end the linear loop instead of looping forever, and a drifted source
closes the session permanently.

```{r}
denied <- cap_agent_session(
  orders,
  budget = 500,
  policy = cap_policy(max_followup_budget = 100)
)
cap_agent_run(denied, function(prompt) {
  list(
    claims = list(),
    evidence = list(),
    warnings = list(),
    requests = list(list(
      fieldId = "f1:table@sample#k10",
      reason = "Too expensive for this budget."
    ))
  )
})
denied$stop_reason
```

## Driving the loop from aisdk

With the suggested [aisdk](https://github.com/YuLab-SMU/aisdk) package
installed, the same session becomes a set of native tools --
`capr_read_digest`, `capr_request_fields`, `capr_submit_claims`, and
`capr_session_status` -- and the aisdk agent runtime handles the tool loop.
The provider credentials and the model call belong entirely to aisdk; capR
only supplies the evidence tools.

```{r, eval = FALSE}
library(aisdk)

session <- cap_agent_session(
  orders,
  question = "How many orders are there, and what do they look like?",
  budget = 500,
  policy = cap_policy(max_followup_budget = 300)
)

analyst <- cap_aisdk_agent(session, model = "claude-sonnet-5")
result <- analyst$run("How many orders are there?")

# Or register the tools on any aisdk loop yourself:
tools <- cap_aisdk_tools(session)
```

However the model behaves, the host-side invariants hold: it can only cite
disclosed field ids, only request fields listed in
`<available_on_request>`, and every extra disclosure is decided by
`cap_gate()` under the session policy -- never by the model.

Model Context Protocol exposure is deliberately deferred: aisdk currently
ships no MCP server, and ADR-0006 records the alternative (posit-dev
mcptools over the same provider-neutral tool layer) for when a host needs
one.
