---
title: "Getting started with capR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with capR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

In the next five minutes you will compile a data frame into a *digest* —
the bounded evidence pack a model is allowed to read — then learn to read
one yourself, and mechanically check a model-style answer against it.
capR never contacts a model; every code block here runs offline.

## The setup

Here is a table you would never paste into a prompt: it contains live
API tokens.

```{r}
library(capR)

orders <- data.frame(
  order_id = c("A001", "A002"),
  amount = c(12.5, 19),
  api_token = c("sk_test_123", "sk_test_456"),
  check.names = FALSE
)
```

## Compile it into a digest

```{r}
digest <- cap_digest(orders, budget = 500, label = "orders")
print(digest)
```

The print method already tells the story: some fields were selected, some
were not, and disclosure was metered against your budget. Now look at the
exact text a model would receive:

```{r}
cat(digest$text)
```

## How to read a digest

Four things to notice, top to bottom:

1. **The header line** states the format, the source fingerprint, the
   tokenizer used for cost accounting, and `budget=used/requested` — the
   meter, up front.
2. **`<field>` blocks are the only facts.** Each has a stable, citable id
   such as `f1:table@shape#base`. Anything a model later claims must cite
   these ids.
3. **The secrets are already gone.** The `api_token` column is *named* in
   the column summary, but its values were masked by the redactor before
   the renderer ever saw them:

```{r}
stopifnot(!grepl("sk_test_123", digest$text, fixed = TRUE))
```

4. **`<available_on_request>` is the honest boundary.** Sample rows exist
   but were not disclosed; the line advertises their id, execution class,
   and estimated cost. A model may *ask* for them — it cannot take them.

The machine-readable twin of the text is the manifest, which records every
field considered, selected or rejected, and why:

```{r}
vapply(
  Filter(function(row) row$selected, digest$manifest$fields),
  `[[`, character(1), "fieldId"
)
```

## Check an answer instead of trusting it

Suppose the model replies with a claim and a citation. Validation is a
mechanical cross-check against the manifest and the digest text — not a
judgment call:

```{r}
good <- cap_validate_response(digest, list(
  claims = list(list(
    id = "claim-1",
    text = "The table has two rows and three columns.",
    evidence = list("f1:table@shape#base")
  )),
  evidence = list(),
  warnings = list(),
  requests = list()
))
stopifnot(good$ok)
```

Now the same claim citing a field that was never disclosed:

```{r}
bad <- cap_validate_response(digest, list(
  claims = list(list(
    id = "claim-1",
    text = "The third sample row shows a refund.",
    evidence = list("f1:table@sample#k10")
  )),
  evidence = list(),
  warnings = list(),
  requests = list()
))
bad$ok
bad$errors[[1L]]$code
```

`evidence_rejected_field`: the field exists in the manifest but was never
selected into the digest, so citing it is invalid. Hallucinated citations
fail the same way, deterministically.

## Where to go next

- **Follow-up disclosure** — when the model *requests* the sample field,
  `cap_gate()` decides under your policy and budget, and only approvals
  become patches: `vignette("response-followup")`.
- **Let a model drive the whole loop** with an audit transcript:
  `vignette("agentic-workflow")`.
- **The full story in Chinese**, from policy to artifacts on disk:
  `vignette("quickstart")`.
- **The mental model in one page**: [Concepts](https://github.com/xiayh0107/capR/blob/main/docs/concepts.md).
