---
title: "Strategy plugins: planners, tokenizers, and options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Strategy plugins: planners, tokenizers, and options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

capR's pipeline stays deterministic, but two of its decisions are now
pluggable seams, following the adapter-registry pattern: the **planner**
(which eligible fields are ranked into the budget) and the **tokenizer**
(how rendered output is costed). The seams open ordering and accounting
only — eligibility filtering, execution authorization, the budget commit
walk, redaction ordering, and every gate rule stay inside the runtime, so a
custom strategy cannot express a policy violation.

Defaults are untouched: with `planner = NULL` and `tokenizer = NULL` (or the
built-in ids), artifacts are byte-identical to previous releases, and the
conformance fixtures always run the built-ins.

```{r}
library(capR)
scores <- data.frame(
  sample_id = c("S1", "S2", "S3"),
  qc_score = c(0.91, 0.87, 0.95)
)
```

## A custom planner

A planner supplies one function: `rank(candidates, question, policy)`,
returning a permutation of the eligible candidate indices, most preferred
first. Its id is stamped into `plan$planner` and the resolution sidecar.

```{r}
cost_first <- cap_planner(
  id = "cheapest-first-v1",
  version = "1.0.0",
  rank = function(candidates, question, policy) {
    order(vapply(candidates, `[[`, integer(1), "estimated_cost"))
  }
)
digest <- cap_digest(scores, budget = 130, planner = cost_first)
digest$artifact$plan$planner
digest$provenance$strategies
```

Invalid rankings fail typed instead of corrupting selection:

```{r, error = TRUE}
broken <- cap_planner(
  id = "broken-v1",
  version = "1.0.0",
  rank = function(candidates, question, policy) rep(1L, length(candidates))
)
try(cap_digest(scores, budget = 130, planner = broken))
```

## A custom tokenizer

The built-in `heuristic_v1` counts roughly four characters per token, which
undercounts CJK text. A tokenizer replaces budget accounting wholesale and
is stamped into the digest header, every manifest row, and the sidecar, so
artifacts remain self-describing.

```{r}
cjk_aware <- cap_tokenizer(
  id = "cjk-aware-v1",
  version = "1.0.0",
  count = function(rendered, field_id) {
    wide <- nchar(gsub("[\\x01-\\x7F]", "", rendered, perl = TRUE))
    ascii <- nchar(rendered, type = "chars") - wide
    as.integer(wide + ceiling(ascii / 4))
  }
)
digest <- cap_digest(scores, budget = 800, tokenizer = cjk_aware)
strsplit(digest$text, "\n")[[1]][[1]]
digest$manifest$budget$tokenizer
```

Follow-up patches are pinned to the digest's accounting: `cap_patch()`
refuses a tokenizer whose id differs from `manifest$budget$tokenizer`, so a
digest and its patches are always costed by one accountant.

Registries mirror the adapter registry for reuse across calls:

```{r}
cap_register_tokenizer(cjk_aware)
cap_list_tokenizers()
digest <- cap_digest(scores, tokenizer = "cjk-aware-v1")
cap_unregister_tokenizer("cjk-aware-v1")
```

## Calibrating estimated costs

`estimatedCost` literals are part of an adapter's identity, so capR never
rewrites them at runtime. `cap_calibrate_costs()` measures what each
field/level actually costs so an adapter author can recost deliberately:

```{r}
cap_calibrate_costs(scores)
```

## Package options

Options configure default magnitudes only — never permissions, strategies,
or redaction (see `?capR` for the full table and the rationale):

```{r}
options(capr.default_budget = 300L)
cap_digest(scores)$manifest$budget$requested
options(capr.default_budget = NULL)
```

`cap_run_fixtures()` clears every `capr.*` option for the duration of its
checks, so conformance evidence can never be skewed by session state.
